---
title: "Crosswalking Global Coral Microbiome Data (GCMP) into the Genomic Observatories Metadatabase (GEOME)"
output: html_notebook
---

# Introduction

This notebook documents the code used to translate the Global Coral Microbiome Project's (GCMP) collated metadata (provided by Jesse Zaneveld) into a format that GEOME can ingest. We are hoping to share the GCMP metadata on GEOME and for this to be a use-case to help improve GEOME's ability to store microbiome metadata.

The metadata provided by Jesse are in `GCMP_EMP_map_r29.txt`. The template to which we want to write the data is in `Global Coral Microbiome Project.xlsx`

# Setup

```{r setup}
library(tidyverse)
library(lubridate)
library(xlsx)

```

# Importing the data

```{r}

y_or_n <- function(x){
  x == "y"
}

df <- read_tsv(file = "GCMP_EMP_map_r29.txt", na = c("","NA","na","Not Applicable"),
              col_types = cols(
                date = col_date(format = "%m/%d/%y"),
                pcrtubeno = "c",
                depth = "n",
                temperature = "n",
                surface_temperature = "n",
                colony_height_m = "n",
                colony_width1 = "n",
                colony_width2 = "n"
              )) %>% 
          #this optionally changes all columns that start with "binary" to logical (T/F) instead of y/n
          #mutate(across(starts_with("binary"), function(x){x == "y"}
          #              )) 
          # you may want to continue cleaning the data in this import pipeline statement, 
          # like changing other y/n columns to logical                    
```

# Clean the Data


# Split the data into GEOME tables

Now you'll need to split the columns into the four tables that GEOME uses: events, samples, tissues, and diagnostics  

You can follow what I specified in GCMP_EMP_map_r29_EDC.xlsx (green = events, blue = samples, yellow = tissues, orange = diagnostics)

Still, some decisions need to be made about what gets included in GEOME or not. XXXXX means I think it 
shouldn't be included, ???? means I don't know you should ask or find a darwinCore term, and then I made no decisions
past family (column CU)

```{r}
events <- df %>% select(collection_id,date,reef_name, collection_time, latitude, longitude, substrate, depth)
samples <- df %>% blah.




```

# Rename and convert

Now you need to rename and/or convert any data into GEOME format

```{r}
events <- events %>% rename(eventID = collection_id,
                            locationRemarks = reef_name,
                            timeOfDay = collection_time,
                            decimalLongitude = longitude,
                            decimalLatitude = latitude,
                            substratum = substrate,
                            maximumDepthInMeters = depth) %>% mutate(collectionDay = day(date),
                                                                     collectionMonth = month(date),
                                                                     collectionYear = year(date)) %>%
              select(-date)



```

# Deduplicate

Here is where you'll make the data more efficient, into [First Normal Form](https://en.wikipedia.org/wiki/First_normal_form), 
by removing duplicate entries in the events and samples tables.

Currently consulting with John Deck about moving substratum and depth to the Sample table, rather than events...

```{r}
dim(events)
events <- events %>% distinct()

dim(events)

```


# Split into Expeditions

There were 11 or 14 GCMP expeditions, and we need to split these concatenated data back into their constituent expeditions.

```{r}
E1 <- df %>% mutate(expedition = str_match(SampleID,"10895\\.(E\\d+)")[,2], .before = sample_name_backup) %>% 
        filter(expedition == "E1")
  

```

# Validate

You'll want to use dplyr left_join() here and some sort of diff function to see if you can exactly reconstruct the dataset from these de-duplicated tables



# Write to Excel Template

You'll want to use the `write.xlsx()` function in the xlsx package to write each table to the `Global Coral Microbiome Project.xlsx` GEOME template in the working directory.